An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-12 EconGrowth.Rnw
Added a printCDParamsTable function and used it in the appendix. Cleaned up the appendix quite a bit. Added some further code cleanup.

\section*{To-do List}
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item Heun: Add setup variables for figure height and width where not already included.
\item Heun: Deal with missing covariance numbers when columns have unequal length.
\item Heun: Add single factor fits
      \begin{itemize}
      \item correct sizes for parameter graphs for each country
      \item \st{parameter graphs ($\lambda$ and $m$) for each country}
      \item \st{parameter tables for each country in the appendix}
      \item \st{GDP comparison graph}
      \item \st{AIC rows in the AIC table}
      \end{itemize}
\item Heun: Add CES production function fits
      \begin{itemize}
      \item GDP comparison graph
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Add Linex production function fits
      \begin{itemize}
      \item GDP comparison graph
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email on 4 Feb 2013 but have not heard a response.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item \st{Heun: Ask Martin if he can supply an ``Economics'' affiliation. Will probably make a stronger paper.}
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item \st{Heun: move tables to an appendix at the end.}
\item \st{Heun: Add acknowledgements (Dad, Pruim)}
\item \st{Heun: Add $u$ parameter table and graph}
\item \st{Heun: Add covariance metrics.}
\end{itemize}

@@

###############################
# Prints a covariance table for countryAbbrev
##
printCovarTable <- function(countryAbbrev){
  print(covarianceTable(countryAbbrev), 
                      caption.placement="top", 
                      sanitize.rownames.function = identity,
                      sanitize.colnames.function = identity,
                      size="\\tiny",
                      table.placement="H")
}
@

@@

Figure \ref{fig:Factors_Lattice_Graph} shows historical data, including GDP ($y$), capital stock ($k$), labor ($l$), and the various energy types ($q$, $x$, and $u$). Factors of production track each other closely, and large covariances among these variables can be seen in Tables \ref{tab:Covariance_US} through \ref{tab:Covariance_ZM}.

<<Factors_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="Historical data. Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).">>=
createFactorsLatticeGraph <- function(){
  dataTable <- loadData("All")
  graphType <- "b" #b is for both line and symbol
  lineTypes <- c(0, 1, 5, 2, 4, 1) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(0, 2, 2, 1, 1, 1) #line widths. 0 means no line.
  colors <- c("black", "black", "black", "red", "blue", "darkorange") #line and symbol colors
  symbols <- c(1, NA, NA, NA, NA, NA)  #NA gives no symbol.
  graph <- xyplot(iGDP+iCapStk+iLabor+iQ+iX+iU ~ Year | Country, data=dataTable,
                  type = graphType,
                  index.cond = list(countryOrderForGraphs), #orders the panels.
                  layout = ninePanelLayoutSpec, 
                  strip = strip.custom(factor.levels=countryNamesAlph, bg="white"), #Sets strip parameters
                  as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                  lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                  pch = symbols, col.symbol = colors, #Controls symbol parameters
                  key=list(text=list(c("$y$", "$k$", "$l$", "$q$", "$x$", "$u$")),
                           type=graphType,
                           cex=keyTextSize, #controls size of text in the key
                           lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                           pch=symbols, col=colors, #controls symbol (plot characters) types
                           columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                  scales=list(cex=scaleTextSize, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              y=list(relation="free", #allows each axis to be different
                                     at=c(1,2,3,4,5,6,7,8,9,10) #tick mark locations
                                     )
                              ),
                  ylim=list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)), #y axis limits
                  xlab="", ylab="Indexed (1980=1 or 1991=1)") #axis labels
  return(graph)
}
createFactorsLatticeGraph()
@

@@

###############################
# Prints a covariance table for countryAbbrev
##
printSingleFactorParamsTable <- function(factor){
  print(singleFactorParamsTable(factor),
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny", 
      table.placement="H")
}

@@

@@

<<CES function, eval=FALSE>>=
cesData <- function(countryName, energyType){
  energyColumnName <- paste("i", energyType, sep="")
  # Load the data that we need.
  dataTable <- loadData(countryName)

@@

The following tables provide tabular results for single-factor models.

<<SF_Params_Tables, results='asis'>>=
#I wish I could say this, but I get bogus K [1]” strings appearing in the LaTeX output.
#lapply(factors, printSingleFactorParamsTable)
printSingleFactorParamsTable("K")
printSingleFactorParamsTable("L")
printSingleFactorParamsTable("Q")
printSingleFactorParamsTable("X")
printSingleFactorParamsTable("U")
@

@@

Table \ref{tab:CD_Parameters_No_Energy} gives the parameters for the Cobb-Douglas model without energy.

<<CD_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableNoEnergy(), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny", 
      table.placement="H")
@

The following tables provide tabular results for Cobb-Douglas models with various energy terms.

<<CDq_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("Q"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@

<<CDx_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("X"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@

<<CDu_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("U"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@

@@

<<AIC_Table_Functions>>=
###############################
# Creates an xtable object that holds the AIC values for each parameter estimation that we include.
##
createAICTable <- function(){
  # Single-factor with K
  sfKModels <- lapply(countryAbbrevs, singleFactorModel, factor="K")
  aicSFk <- data.frame(lapply(sfKModels, AIC))
  rownames(aicSFk) <- "SF$k$"
  # Single-factor with L
  sfLModels <- lapply(countryAbbrevs, singleFactorModel, factor="L")
  aicSFl <- data.frame(lapply(sfLModels, AIC))
  rownames(aicSFl) <- "SF$l$"
  # Single-factor with Q
  sfQModels <- lapply(countryAbbrevs, singleFactorModel, factor="Q")
  aicSFq <- data.frame(lapply(sfQModels, AIC))
  rownames(aicSFq) <- "SF$q$"
  # Single-factor with X
  sfXModels <- lapply(countryAbbrevs, singleFactorModel, factor="X")
  aicSFx <- data.frame(lapply(sfXModels, AIC))
  rownames(aicSFx) <- "SF$x$"
  # Single-factor with U
  aicSFu <- cbind(US=AIC(singleFactorModel("US", "U")), 
                  UK=AIC(singleFactorModel("UK", "U")), 
                  JP=AIC(singleFactorModel("JP", "U")),
                  CN=NA, ZA=NA, SA=NA, IR=NA, TZ=NA, ZM=NA) #No U data for these countries.
  rownames(aicSFu) <- "SF$u$"
  # Cobb-Douglas models
  cdModels <- lapply(countryAbbrevs, cobbDouglasModel)
  aicCD <- data.frame(lapply(cdModels, AIC))
  rownames(aicCD) <- "CD"
  # Cobb-Douglas with Q
  cdQModels <- lapply(countryAbbrevs, cobbDouglasModel, energyType="Q")
  aicCDq <- data.frame(lapply(cdQModels, AIC))
  rownames(aicCDq) <- "CD$q$"
  # Cobb-Douglas with X
  cdXModels <- lapply(countryAbbrevs, cobbDouglasModel, energyType="X")
  aicCDx <- data.frame(lapply(cdXModels, AIC))
  rownames(aicCDx) <- "CD$x$"
  # Cobb-Douglas with U
  cdUModels <- c(cobbDouglasModel("US", "U"), cobbDouglasModel("UK", "U"), cobbDouglasModel("JP", "U"))
  aicCDu <- cbind(US=AIC(cobbDouglasModel("US", "U")), 
                  UK=AIC(cobbDouglasModel("UK", "U")), 
                  JP=AIC(cobbDouglasModel("JP", "U")),
                  CN=NA, ZA=NA, SA=NA, IR=NA, TZ=NA, ZM=NA) #No U data for these countries.
  rownames(aicCDu) <- "CD$u$"
  # Create the data.frame table
  out <- rbind(aicSFk, aicSFl, aicSFq, aicSFx, aicSFu, aicCD, aicCDq, aicCDx, aicCDu) 
  # Create the xtable with caption and label
  out <- xtable(out, caption="AIC values for all models.", label="tab:AICTable", digits=1)
  return(out)
}
@

>>>>

\section*{To-do List}
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item Heun: Add CES production function fits
      \begin{itemize}
      \item GDP comparison graph
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Add Linex production function fits
      \begin{itemize}
      \item GDP comparison graph
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Deal with missing covariance numbers when columns have unequal length.
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email on 4 Feb 2013 but have not heard a response.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item \st{Heun: Ask Martin if he can supply an ``Economics'' affiliation. Will probably make a stronger paper.}
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item \st{Heun: move tables to an appendix at the end.}
\item \st{Heun: Add acknowledgements (Dad, Pruim)}
\item \st{Heun: Add $u$ parameter table and graph}
\item \st{Heun: Add covariance metrics.}
\item Heun: Add single factor fits
      \begin{itemize}
      \item \st{correct sizes for parameter graphs for each country}
      \item \st{parameter graphs ($\lambda$ and $m$) for each country}
      \item \st{parameter tables for each country in the appendix}
      \item \st{GDP comparison graph}
      \item \st{AIC rows in the AIC table}
      \end{itemize}
\item \st{Heun: Add setup variables for figure height and width where not already included.}
\end{itemize}

@@

###############################
# Prints a covariance table for countryAbbrev
##
printCovarTable <- function(countryAbbrev){
  print(covarianceTable(countryAbbrev), 
                      caption.placement="top", 
                      sanitize.rownames.function = identity,
                      sanitize.colnames.function = identity,
                      size="\\tiny",
                      table.placement="H")
}

####################################
# Creates a graph that displays all of the factors of production for all countries.
##
createHistoricalLatticeGraph <- function(){
  dataTable <- loadData("All")
  graphType <- "b" #b is for both line and symbol
  lineTypes <- c(0, 1, 5, 2, 4, 1) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(0, 2, 2, 1, 1, 1) #line widths. 0 means no line.
  colors <- c("black", "black", "black", "red", "blue", "darkorange") #line and symbol colors
  symbols <- c(1, NA, NA, NA, NA, NA)  #NA gives no symbol.
  graph <- xyplot(iGDP+iCapStk+iLabor+iQ+iX+iU ~ Year | Country, data=dataTable,
                  type = graphType,
                  index.cond = list(countryOrderForGraphs), #orders the panels.
                  layout = ninePanelLayoutSpec, 
                  strip = strip.custom(factor.levels=countryNamesAlph, bg="white"), #Sets strip parameters
                  as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                  lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                  pch = symbols, col.symbol = colors, #Controls symbol parameters
                  key=list(text=list(c("$y$", "$k$", "$l$", "$q$", "$x$", "$u$")),
                           type=graphType,
                           cex=keyTextSize, #controls size of text in the key
                           lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                           pch=symbols, col=colors, #controls symbol (plot characters) types
                           columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                  scales=list(cex=scaleTextSize, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              y=list(relation="free", #allows each axis to be different
                                     at=c(1,2,3,4,5,6,7,8,9,10) #tick mark locations
                                     )
                              ),
                  ylim=list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)), #y axis limits
                  xlab="", ylab="Indexed (1980=1 or 1991=1)") #axis labels
  return(graph)
}
@

@@

Figure \ref{fig:Historical_Lattice_Graph} shows historical data, including GDP ($y$), capital stock ($k$), labor ($l$), and the various energy types ($q$, $x$, and $u$). Factors of production track each other closely, and large covariances among these variables can be seen in Tables \ref{tab:Covariance_US} through \ref{tab:Covariance_ZM}.

<<HIstorical_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="Historical data. Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).">>=
createHistoricalLatticeGraph()
@

@@

###############################
# Prints a covariance table for the given factor of production
##
printSFParamsTable <- function(factor){
  print(singleFactorParamsTable(factor),
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny", 
      table.placement="H")
}

@@

############################
# Prints a table with parameters from a Cobb-Douglas model for the given energyType. 
# Leave energyType blank to print a table for Cobb-Douglas without energy.
printCDParamsTable <- function(energyType){
  if (missing(energyType)){
    print(cobbDouglasParamsTableNoEnergy(), 
        caption.placement="top", 
        sanitize.colnames.function = identity, 
        size="\\tiny", 
        table.placement="H")
  } else {
    print(cobbDouglasParamsTableWithEnergy(energyType), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
  }
}

@@

<<CES functions, eval=FALSE>>=
cesData <- function(countryName, energyType){
  energyColumnName <- paste("i", energyType, sep="")
  # Load the data that we need.
  dataTable <- loadData(countryName)

@@

The following tables provide results for parameters in single-factor models.

<<SF_Params_Tables, results='asis'>>=
#I wish I could say this, but I get bogus K [1]” strings appearing in the LaTeX output.
#lapply(factors, printSingleFactorParamsTable)
printSFParamsTable("K")
printSFParamsTable("L")
printSFParamsTable("Q")
printSFParamsTable("X")
printSFParamsTable("U")
@

@@

The following tables provide results for parameters in Cobb-Douglas models with and without energy terms.

<<CD_Params_Table, results='asis'>>=
printCDParamsTable()    #Without energy
printCDParamsTable("Q")
printCDParamsTable("X")
printCDParamsTable("U")
@

\subsection{CES Models}

The following tables provide results for parameters in CES models.

\subsection{LINEX Models}

The following tables provide results for parameters in LINEX models.

@@

<<AIC_Table_Functions>>=
###############################
# Creates an xtable object that holds the AIC values for each parameter estimation that we include.
##
createAICTable <- function(){
  ######################
  # Single-factor models
  ######################
  # Single-factor with K
  sfKModels <- lapply(countryAbbrevs, singleFactorModel, factor="K")
  aicSFk <- data.frame(lapply(sfKModels, AIC))
  rownames(aicSFk) <- "SF$k$"
  # Single-factor with L
  sfLModels <- lapply(countryAbbrevs, singleFactorModel, factor="L")
  aicSFl <- data.frame(lapply(sfLModels, AIC))
  rownames(aicSFl) <- "SF$l$"
  # Single-factor with Q
  sfQModels <- lapply(countryAbbrevs, singleFactorModel, factor="Q")
  aicSFq <- data.frame(lapply(sfQModels, AIC))
  rownames(aicSFq) <- "SF$q$"
  # Single-factor with X
  sfXModels <- lapply(countryAbbrevs, singleFactorModel, factor="X")
  aicSFx <- data.frame(lapply(sfXModels, AIC))
  rownames(aicSFx) <- "SF$x$"
  # Single-factor with U
  aicSFu <- cbind(US=AIC(singleFactorModel("US", "U")), 
                  UK=AIC(singleFactorModel("UK", "U")), 
                  JP=AIC(singleFactorModel("JP", "U")),
                  CN=NA, ZA=NA, SA=NA, IR=NA, TZ=NA, ZM=NA) #No U data for these countries.
  rownames(aicSFu) <- "SF$u$"
  ######################
  # Cobb-Douglas models
  ######################
  cdModels <- lapply(countryAbbrevs, cobbDouglasModel)
  aicCD <- data.frame(lapply(cdModels, AIC))
  rownames(aicCD) <- "CD"
  # Cobb-Douglas with Q
  cdQModels <- lapply(countryAbbrevs, cobbDouglasModel, energyType="Q")
  aicCDq <- data.frame(lapply(cdQModels, AIC))
  rownames(aicCDq) <- "CD$q$"
  # Cobb-Douglas with X
  cdXModels <- lapply(countryAbbrevs, cobbDouglasModel, energyType="X")
  aicCDx <- data.frame(lapply(cdXModels, AIC))
  rownames(aicCDx) <- "CD$x$"
  # Cobb-Douglas with U
  cdUModels <- c(cobbDouglasModel("US", "U"), cobbDouglasModel("UK", "U"), cobbDouglasModel("JP", "U"))
  aicCDu <- cbind(US=AIC(cobbDouglasModel("US", "U")), 
                  UK=AIC(cobbDouglasModel("UK", "U")), 
                  JP=AIC(cobbDouglasModel("JP", "U")),
                  CN=NA, ZA=NA, SA=NA, IR=NA, TZ=NA, ZM=NA) #No U data for these countries.
  rownames(aicCDu) <- "CD$u$"
  ######################
  # CES models
  ######################

  ######################
  # LINEX models
  ######################

  # Create the data.frame table
  out <- rbind(aicSFk, aicSFl, aicSFq, aicSFx, aicSFu, aicCD, aicCDq, aicCDx, aicCDu) 
  # Create the xtable with caption and label
  out <- xtable(out, caption="AIC values for all models.", label="tab:AICTable", digits=1)
  return(out)
}
@
